Drought-sensitive trees survive on deep water access in a tropical forest community

### 2020-08-01 manuscript.rmd
Edited first Discussion para and added two

\
We test an overarching hypothesis that interspecific variation in effective rooting depths of trees aligns species on a fast-slow trade-off axis. Specifically, (i) dynamics of soil water potential at depth does not cross critical plant thresholds even during multi-annual droughts because of high total annual precipitation to evapotranspiration ratio (not sure if this should be stated as a hypothesis or just an explanation of the results); (ii) trees with deeper ERD show 'fast' hydraulic traits: higher maximum stem hydraulic conductance ($K_{max}$), but greater xylem vulnerability to embolism ($\Psi_{88,~stem}$), higher $\Psi_{tlp}$ and narrower hydraulic safety margins ($\Psi_{min} - \Psi_{88,~stem}$); and finally, (iii) deeper ERD species have lower mortality rates due to low realised risk of hydraulic failure. To our knowledge this is the first study to demonstrate a mechanistic link between trees' realised drought experience, above-ground hydraulic architecture and mortality outcomes for a species-rich forest.

@@

### Model parameterization
\
To parameterise the model, we identified eleven parameters relevant for Eq. \@ref(eq:water-balance), determined their ranges based on literature for the study site or for the tropics (Table \@ref(tab:param)), and sampled 5000 ensembles from this parameter space with Latin Hypercube Sampling (LHS). In the Conrad catchment soil hydraulic conductivity declined by depth based on existing data for the depths of 12.5 and 60 cm (@Kinner:2004; (Table \@ref(tab:param)). We generated 5000 LHS samples within the 95% CI for the two depths and assumed a linear decline between each paired sample, while for depths < 12.5 cm and > 60 cm, we assumed the same value as that for depths 12.5 and 60 cm, respectively (Fig. S\@ref(fig:ksat-profiles)). For all runs, instead of ELM's default soil texture-based soil water retention curve, we used a curve that we derived from existing data on volumetric water content ($\theta$) vs. $\Psi_{soil}$ for depths 15, 40 and 100 cm [@Kupers:2019]. A Campbell (1974) empirical equation was fitted for all depths pooled together, as exploratory analyses indicated differences among depths were nominal:

@@

# Discussion

Our analyses revealed that trees with deeper effective rooting depths (ERD) showed acquisitive hydraulic architecture, with higher maximum stem hydraulic conductance, but greater xylem vulnerability to embolism and narrower hydraulic safety margins compared to shallow ERD species. Yet, deeper ERD species showed lower mortality rates over the 33 year record (1982-2015). Analysis of time spent below ($\Psi_{crit}$) in the soil zone of ERD showed that this was likely due to low realised risk of hydraulic failure for deeper ERD species. We thus show, for the first time, that interspecific variation in trees' effective rooting depths aligns species on a 'fast-slow' water economic spectrum by demonstrating a link between trees' realised drought experience, hydraulic architecture and long-term mortality outcomes. On the 'fast' end of the spectrum were species with hydraulically efficient, but risky suit of hydraulic traits. These were associated with, and likely optimised for, an environment of reliable and abundant source of water at depth, which actualised to a lower associated mortality-risk. At the slow end were species with inefficient, but safer xylem, associated with a shallow soil- water environment of high variability.

## Higher xylem embolism vulnerability of species with Deeper Effective Rooting Depths

-- trade-offs among differential investment in hydraulic efficiency and safety
here links with trade-off in water-stress vertically belowground
-- unexplored (hydraulic capacitance, stomatal control, photosynthetic capacity, high temperature tolerance)

fast-slow axis is matched by presence of an efficiency-safety trade-off between high water conductance (high $K_{max}$) and cavitation-resistant xylem (low $\Psi_{50}$), although community-wide examples are rare [@Hacke:2006;@Gleason:2016;]. Low $\Psi_{50}$ is in turn associated with low $\Psi_{tlp}$ and both are broadly associated with low water availability[@Choat:2012; @Bartlett:2012]. Slow strategy is associated with larger safety-margin ($\Psi_{min} - \Psi_{88,~stem}$) in two angiosperm systems. High $K_{max}$ is associated with photosynthetic rates with faster growth potential.

## Lower mortality of species with deeper Effective Rooting Depths

Mortality could be immediate or may result after decades of drought occurrence[@Cailleret:2016].

## Limitations & Future directions

Our modeling framework had a few limitations. 
- The inverse model for ERD was parametarised on 5-yr growth rates, which vastly increased the degree of freedom to fit the model. Future studies should use high-frequency growth time series such as those obtained from dendrometer bands. To infer growth from such diameter time series, however, would especially require accounting for diameter changes due to hydrostatic flexing [@Chitra-Tarak:2015].

- Future studies should investigate seasonality in water-sourcing depths investigated.

- Although we attempted a coarse estimate of species leaf cover seasonality by combining leaf-fall data from litter traps and forest-level LAI data for leaf-on periods, more accuarate estimations of leaf-cover seasonality are required, such as those obtained from drone images [@Park.2019]

-To enable estimation of inverse model for the larger community across size classes would need accounting for changing light environments on growth; we minimised this issue by limiting our ERD inversion to large trees of species with max height >= 30m and thus whose canopy's are most likely to be fully exposed to light.

- Future studies would benefit from modeling the complete hydrodynamics of water potential gradients over the soil - plant (roots, stems, leaves) - atmosphere continuum (SPAC) [@Johnson:2018fad; @Christoffersen:2016kd2d]. However, to parameterise such a model at community-scale would need community-scale data on hard traits such as stem and root vulnerability curves; if not, model development would be needed to up-scale data from a few species, for example, via softer traits as we did for leaf vulnerability curves. (Note that In this forest LMA and WSG did not scale with stem vulnerability curves). Water storage (capacitance) should also be considered.

- Deep ERD species may possibly be associated with locations of greater water availability rather than having deep rootedness. A 3-D hydrology model would provide spatially distributed estimates of water dynamics in the whole-soil/rock column and depth to water table to link with individual trees water experience [@Schwantes:2018]. Forest inventory sites would benefit from investing data collection efforts on water availability dynamics by depth and water tables covering spatial heterogeneity so as to better constrain such models.

- Variation in mortality explained by ERD is low. Other causes.... Based on a year-long study, lightening explains ~4.5 % of mortality for trees >=10 cm dbh and ~ 40% for trees >=60 cm dbh.

-Forest is realtively stable over 35 years. Future studies could investigate if ongoing change in the duration, frequency and intensity of droughts entails sufficient exhaustion of deep water availablity so as to cause mortality of deep ERD species.

## Conclusions:

>>>>

\
We test an overarching hypothesis that interspecific variation in effective rooting depths of trees aligns species on a fast-slow economic spectrum optimised for increasingly reliable water availability by depth such that (i) trees with deeper ERD have hydraulic traits associated with rapid water transport but cavitation-vulnerable xylem: higher maximum stem hydraulic conductance ($K_{max}$), but greater xylem vulnerability to embolism ($\Psi_{88,~stem}$), higher $\Psi_{tlp}$ and narrower hydraulic safety margins ($\Psi_{min} - \Psi_{88,~stem}$), whereas shallower ERD species are associated with less efficient but safer hydraulic traits; and that, (ii) deeper ERD species have lower mortality rates due to low realised risk of hydraulic failure, whereas shallower ERD species may have higher mortality rates as a result of higher variability in shallow soil water availability. To our knowledge this is the first study to demonstrate a mechanistic link between trees' realised drought experience, above-ground hydraulic architecture and long-term mortality outcomes for a species-rich forest.

@@

### Model parameterization
\
To parameterise the model, we identified eleven parameters relevant for Eq. \@ref(eq:water-balance), determined their ranges based on literature for the study site or for the tropics (Table \@ref(tab:param)), and sampled 5000 ensembles from this parameter space with Latin Hypercube Sampling (LHS). In the Conrad catchment soil hydraulic conductivity declined by depth based on existing data for the depths of 12.5 and 60 cm (@Kinner:2004; (Table \@ref(tab:param)). We generated 5000 LHS samples within the 95% CI for the two depths and assumed a linear decline between each paired sample, while for depths < 12.5 cm and > 60 cm, we assumed the same value as that for depths 12.5 and 60 cm, respectively (Fig. S\@ref(fig:ksat-profiles)). For all runs, instead of ELM's default soil texture-based soil water retention curve, we used a curve that we derived from existing data on volumetric water content ($\theta$) vs. $\Psi_{soil}$ for depths 15, 40 and 100 cm [@Kupers:2019]. A Campbell @Campbell.1974 empirical equation was fitted for all depths pooled together, as exploratory analyses indicated differences among depths were nominal:

@@

# Discussion
\
Our analyses revealed that trees with deeper effective rooting depths (ERD) were associated higher hydraulic efficiency, but not safety: higher maximum stem hydraulic conductance, but greater xylem vulnerability to embolism and narrower hydraulic safety margins compared to shallow ERD species. Yet, deeper ERD species showed lower mortality rates than shallower ERD species over the 33 year record (1982-2015). Analysis of time spent below ($\Psi_{crit}$) in the soil zone matching ERD $\Psi_{soil,~z = ERD}$ showed that this was likely due to decreasing realised risk of hydraulic failure with depth. We thus show, for the first time, that interspecific variation in trees' effective rooting depths aligns species on a 'fast-slow' water economic spectrum by demonstrating a link between trees' realised drought experience, hydraulic architecture and long-term mortality outcomes. On the 'fast' end of the spectrum were deep ERD species with hydraulically efficient, but risky suit of traits optimised for an environment of reliable and abundant source of water at depth that actualised to a lower mortality-risk. At the slow end of the spectrum were shallow ERD species with inefficient, but safer xylem, likely optimised for high variability in $\Psi_{soil}$ in the shallow soil layers, resulting in higher variance in mortality rates across censuses as $\Psi_{soil,~z = ERD}$ crossed ERD $\Psi_{crit}$. Mortality rates of shallow ERD species may have been exacerbated by increased drought events during the study period. Previous work shows that half of the extreme dry seasons, estimated as < 1 SD annual water deficit, during 1931-2015 (84 years) occurred during the first 23 years (1982-2005) of the study period. [@Condit.2017]

### Effective rooting depths at tree community level
\
By estimating ERD for large trees of 33 canopy tree species, we make an important advance in modeling effective rooting depths at the community level in species species-rich tropical forests. We draw attention to the availability of data on $\delta^2^H$ on trees and water sources belowground--soil or substrate matrix water in the unsaturated vadoze zone or free water below the water table--as important for model calibration/validation. The first inverse model of ERD (water uptake depth index) developed by Chitra-Tarak et al @Chitra-Tarak:2018 lacked validation as existing methods were inadequate for extracting water bound to the soil and saprolite found in their dry forest study site (Mudumalai National Park, Southern India; Jean Riotte, pers. comm.). Our ERD model was distinctly constrained on species leaf vulnerability (PLC) curves. We benefitted from finding a strong scaling relationship between commonly collected 'soft' traits at the community scale (SLA~lamina~ and Wood Specific Gravity) and parameters of PLC curves for which we collected data on a few species (*n* = 14), a 'hard' trait. This allowed us to scale-up PLC curves at the community scale (*n* = 284 species). Whether such scaling relationships between hard and soft traits hold true for other forests and their wider applicability remains to be seen as data on PLC curves in species-rich, tropical forests are limited, but counter examples exist [see xx, yy, zz (Lawren, could you suggest references?].

\
##### Future Directions for Model Development
\
Our model estimated ERD as a single value, unlike the dynamic ERD modelled by Chitra-Tarak et al @Chitra-Tarak:2018, but future model developments should investigate seasonal dynamics of ERD, calibrated on corresponding calibration data on isotopic water sourcing depths. Our inverse ERD model was parametarised on 5-yearly growth data, with five data points over a 25 year period, which provided large degree of freedom to fit the model. Future studies could better constrain the model with higher resolution growth data such as those from dendrometers. This is especially important in modeling seasonally dynamic ERD. However, trees' stem diameter change is a sum total of not just growth, the addition of wood, but also $\pm$ changes due to hydrostatic stem flexing, whose importance increases with temporal resolution [@Chitra-Tarak.2016; @Chitra-Tarak.2014]. Use of dendrometer data would thus entail accounting for stem hydrostatic flexing [@Zweifel:2018]. In light-limited forests such as BCI, expanding ERD estimates to smaller trees, especially via parametairising over a few growth data points, would likely need model development to account for variation in trees' light environment over time and among trees' of the same size class, an issue we minimised by limiting ERD estimates to canopy trees. To estimate topographic variation in ERD within or across species a distributed hydrological model could be used. We note that data availability on soil water retention curve ($\Psi_{soil}$vs. VWC), hydraulic conductivity ($K_{soil}$) by depth and soil moisture by depth were important for effective calibration of our 1-D hydrological model, and recommend wide-spread data collection of these variables in forested sites as well as water-table levels (not collected in our site) to estimate trees' water environments. Finally, developing an inverse ERD model that accounts for movement of water within the soil-plant-atmosphere-continuum (SPAC) would be a step change from the current statistical models [this study; @Chitra-Tarak:2018]. Although data needs for parametarising such models [@Christoffersen.2016;@Sperry.1998] could be greater (e.g. retention curves for roots), limiting community-wide application, they would improve our understanding of plant physiology [@Johnson:2017].

### Need for drought risk

Plant traits hold clues for the historical environments that shaped their ecological strategy.

## Conclusions
